Seismic Hazard Estimation of Northern Iran Using Smoothed Seismicity

### 2016-07-18 abstract.tex
edits suggested by Naeem

% 
This article presents a seismic hazard assessment for northern Iran, where a smoothed seismicity approach has been used in combination with an updated seismic catalog and a ground motion prediction equation recently found to yield good fit with data. We evaluate the hazard over a geographical area including the seismic zones of Azerbaijan, the Alborz Mountain Range, and Kopeh-Dagh; and for completeness, we also consider the contribution of parts of other neighboring seismic zones that fall within our region of interest. In the chosen approach, seismic events are not assigned to specific faults but are instead assumed to be potential seismogenic sources, and are spatially distributed within regular grid cells of size equal to 0.1\textdegree{} in both latitude and longitude. After performing the corresponding magnitude conversions, we decluster both historical and instrumental seismicity catalogs to obtain earthquake rates based on the number of events within each cell, and smooth the results to account for the uncertainty in the spatial distribution of future earthquakes. Seismicity parameters are computed for each seismic zone separately, and for the entire region of interest as a single uniform seismotectonic region. We also consider the combination ot these two basic models. The results are presented in terms of expected peak ground acceleration maps and hazard curves at select locations, considering exceedance probabilities of 2 and 10 percent in 50 years for rock site conditions. We analyze our results in light of similar estimates available in the literature and offer our perspective on the differences observed. In conclusion, we find our results to be helpful in understanding seismic hazard for northern Iran from a regional perspective, but highlight the fact that different methods may need to be combined to obtain more robust estimates at specific areas of interest.

>>>>

% 
This article presents a seismic hazard assessment for northern Iran, where a smoothed seismicity approach has been used in combination with an updated seismic catalog and a ground motion prediction equation recently found to yield good fit with data. We evaluate the hazard over a geographical area including the seismic zones of Azerbaijan, the Alborz Mountain Range, and Kopeh-Dagh; and for completeness, we also consider the contribution of parts of other neighboring seismic zones that fall within our region of interest. In the chosen approach, seismic events are not assigned to specific faults but are instead assumed to be potential seismogenic sources, and are spatially distributed within regular grid cells of size equal to 0.1\textdegree{} in both latitude and longitude. After performing the corresponding magnitude conversions, we decluster both historical and instrumental seismicity catalogs to obtain earthquake rates based on the number of events within each cell, and smooth the results to account for the uncertainty in the spatial distribution of future earthquakes. Seismicity parameters are computed for each seismic zone separately, and for the entire region of interest as a single uniform seismotectonic region. We also consider the combination of these two basic models. The results are presented in terms of expected peak ground acceleration maps and hazard curves at selected locations, considering exceedance probabilities of 2 and 10 percent in 50 years for rock site conditions. We analyze our results in light of similar estimates available in the literature and offer our perspective on the differences observed. In conclusion, we find our results to be helpful in understanding seismic hazard for northern Iran from a regional perspective, but highlight the fact that different methods may need to be combined to obtain more robust estimates at specific areas of interest.

### 2016-07-18 introduction.tex
edits suggested by Naeem

In the following sections we describe our region of interest in greater detail, review its associated tectonics and seismicity, and provide additional information about the adopted approach. Significant aspects of our work include the composition of the earthquake catalog, including catalog completeness and the conversion of earthquake magnitudes, the selection of an appropriate attenuation relationship for the region, and the estimation of the Gutenberg-Richter seismicity parameters. Our analysis includes seismic hazard curves and maps for peak ground acceleration (PGA) considering exceedance probabilities of 2 and 10 percent in 50 years for rock site conditions for events with moment magnitude $M_w /geq 5$. We analyze our results in light of similar estimates available in the literature and summarize the most relevant conclusions.

>>>>

In the following sections we describe our region of interest in greater detail, review its associated tectonics and seismicity, and provide additional information about the adopted approach. Significant aspects of our work include the composition of the earthquake catalog, including catalog completeness and the conversion of earthquake magnitudes, the selection of an appropriate attenuation relationship for the region, and the estimation of the Gutenberg-Richter seismicity parameters. Our analysis includes seismic hazard curves and maps for peak ground acceleration (PGA) considering exceedance probabilities of 2 and 10 percent in 50 years for rock site conditions for events with moment magnitude $M_w \geq 5$. We analyze our results in light of similar estimates available in the literature and summarize the most relevant conclusions.

### 2016-07-18 results.tex
edits suggested by Naeem

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-08.pdf} 
    \caption{Expected mean peak ground acceleration (PGA) for 2 and 10 percent of probability for earthquake magnitudes exceeding $M_w$ 5 for (a) the five-seismic regions R model and (b) the uniform seismic region U model.}
    \label{fig:pga}
\end{figure*}

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-09.pdf} 
    \caption{Difference between the mean PGA values obtained for the uniform U model results and the corresponding R models for 2 (left) and 10 (right) percent probabilities of exceeding magnitudes $M_w$ 5.}
    \label{fig:pga.diff}
\end{figure*}

@@

Using the data, parameters, and procedures just described, we considered two different hazard models based on whether we use the five distinct seismic hazard regions or the whole area of interest as a uniform seismic hazard region. We call these the R and U models, respectively. Using these models we obtained hazard curves and expected pGA values for two two probabilities of exceedance, at 2 and 10 percent in 50 years for events exceeding magnitudes $M_w \geq 5$. These probabilities correspond to return periods of 475 and 2475 years. For nomenclature purposes, we use the probabilities as a prefix and the magnitude as a suffix on the model type. As an example, we refer to the results for 2 percent probability of exceeding an event magnitude $M_w$ 5 in 50 years using the five-regions model as the 2R5, whereas for the case of 10 percent probability of exceeding $M_w$ 5 when using the uniform model we would use the tag name 10U5.

@@

Fig.~\ref{fig:pga} shows the expected mean PGA values obtained using the selected attenuation relationship as a measured of gravity (g) throughout the region of interest for the different models, 2R5, 10R5, 2U5 and 10U5. As can be expected, the highest PGA values are associated with the 2R5 and 2U5 models, and the lowest with the 10R5 and 10U5 models. In results for the R models shown Fig.~\ref{fig:pga}a, the highest levels of expected ground motion occur in the Kopeh Dagh region (near the border with Turkmenistan), and in a small area half-way between the city of Karaj and the Piranshahr fault, where a small concentration of events can be observed in Fig.~\ref{fig:catalog}b, as well as along the Piranshahr and Morvarid fault lines between the Zagros and Azerbaijan seismic regions (see Fig.~\ref{fig:selected}a for reference). The Azerbaijan seismic zone does not present equally high levels of expected ground motion. In the U models shown in Fig.~\ref{fig:pga}b, on the other hand, the highest levels of expected ground motion are observed precisely in the area corresponding to the Azerbaijan seismic region, especially in the vicinity of the North Tabriz Fault system (see Fig.~\ref{fig:selected} for reference).

@@

As it can be seen in Fig.~\ref{fig:pga.ru.std}, the combined (R,U) model follows more closely the distribution of the region's seismicity shown in Fig.~\ref{fig:catalog} as well as the major faults shown in Fig.~\ref{fig:selected}a, and thus we consider it to be a better assessment of the seismic hazard in the region. It displays well the contribution of the North Tabriz Fault system in the Azerbaijan region as well as the contribution of the Piranshahr and Morvarid faults in the Zagros region and that of the different faults in the Alborz and Kopeh Dagh regions. In Fig.~\ref{fig:pga.ru.std}, the highest PGAs are observed in the 2(R,U)5+STD model results, where the expected accelerations are of the order of 1.4 g to the northwest of city of Tabriz. These maximum values should, nonetheless, be interpreted with caution given the fact that predictions are based on rock-site assumptions and do not include any kind of site effects or other related conditions which may often contribute to mitigate peak accelerations.

@@

We also note that the overall distribution of mean PGA values shown in the middle panels in Fig.~\ref{fig:pga.ru.std} are the ones that best agree with the recent study by \citet{Khodaverdian_2016_BSSA}. These estimates also agree, on average, with other previous studies. For the north-west seismic regions of Azerbaijan and Zagros, for instance, \citet{Tavakoli1999} predicts maximum mean PGA (10\% in 50 years) values of 0.45 g near the North Tabriz Fault and the North Tehran Fault zones, where we find equivalent values to be between 0.35 and 0.5 g in the 10(R,U)5 model. Similarly, for the particular case of the city of Tabriz, \citet{Vafaie2011} obtained slightly higher expected accelerations on the order of 0.30--0.90 and 0.20--0.65 g for 2 and 10\%, respectively, where our mean values are about 0.53 and 0.36 g for the same corresponding probabilities. In the north-central region of Alborz, \citet{Ghodrati2003} and \citet{Boostan2015} predict PGA values of the order 0.27--0.46 g and 0.42--0.48 for Tehran, where we found equivalent results to be between 0.50 and 0.30 g, for 2 and 10\% respectively. The latter estimate is in good agreement with \citet{Abdi2013}, who predicts PGA values in the vicinity of Karaj and Tehran to be on the order of 0.27--0.31 g for 10\% probability of exceedance. Results from \citet{Ghodrati2008} for the Gilan province around the city of Rasht are also on the same order of those shown in Fig.~\ref{fig:pga.ru.std}. Finally, for the city of Bojnurd in the northeast province of Khorasan in the Kopeh Dagh seismic region, \citet{Rahgozar2012} reports expected PGA values in the ranges of 0.28--0.49 and 0.15--0.22 g for 2 and 10\%, where we obtain mean values of 0.51 and 0.34 g, respectively.

Table \ref{tab:pga} shows a condensed version of these comparisons between our results and those of other previous studies for selected cities in northern Iran. As mentioned before, overall, our results are close or within the ranges from most studies. The only noticeable exceptions are the results of \citet{Golara2014}, whose expected PGA values are significantly higher the rest, especially in the cities of Tabriz, Karaj, Tehran and Mashhad, where their results are closer to those of our upper (+STD) range of values. Although not included here for brevity, we also computed expected PGA values for a magnitude thresholds $M_w$ 4.5. In that case, our results are closer to those of \citet{Golara2014}, but begin to deviate from the rest. We should note that, in general, the smoothed seismicity method used here tends to lead to lower hazard estimations than those obtained using specific fault definitions (as opposed to point sour\-ces), and thus is sensitive to the appropriate selection of the magnitude threshold.

Following the comparisons shown in Table \ref{tab:pga}, we looked into the more complete picture of the annual probabilities of exceedance as functions of PGA. The results are shown in Fig.~\ref{fig:hazardcurve} for six of the main cities in northern Iran considering the two basic models R5 and U5 and the combined model (R,U)5. In each case we also show the range $\pm 1$ standard deviation (shaded areas). These hazard curves exemplify again the differences between the R and U models, which are more prominent in the northwest cities of Tabriz and Ardabil, where the U models yield higher expected values of PGA for any given annual probability of exceedance. Minor differences are also present in the results of the central cities of Karaj and Tehran, whereas for the cities of Urmia and Mashhad in the Zagros and Kopeh Dagh seismic zones, the R and U models yield similar if not equivalent hazard estimates. This is consistent with our previous observations regarding Figs.~\ref{fig:pga}, \ref{fig:pga.diff} and \ref{fig:pga.ru.std}.

@@

Last, we put additional attention on the results for the capital city of Tehran. We compare hazard curves obtained in this study against those computed by \citet{Ghodrati2003} and \citet{Boostan2015}. Our results are similar to those obtained by \citet{Ghodrati2003} in the range between 0.2 and 0.7 (corresponding to annual probabilities of exceedance between $10^{-5}$ and $10^{-2}$), but the trend indicates we would obtain increasingly higher probabilities associated with PGA values above 0.8 g. Conversely, the trend in the hazard curve from \citet{Boostan2015} suggest the city is to expect significantly higher levels of ground motion at higher probability rates, which is contrary to our findings and those of \citet{Ghodrati2003}.

>>>>

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-08.pdf} 
    \caption{Expected mean peak ground acceleration (PGA) for 2 and 10 percent of probability of exceedance in 50 years for (a) the five-seismic regions R model and (b) the uniform seismic region U model.}
    \label{fig:pga}
\end{figure*}

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-09.pdf} 
    \caption{Difference between the mean PGA values from models U and R for 2 (left) and 10 (right) percent probabilities of exceedance in 50 years.}
    \label{fig:pga.diff}
\end{figure*}

@@

Using the data, parameters, and procedures just described, we considered two different hazard models based on whether we use the five distinct seismic hazard regions or the whole area of interest as a uniform seismic hazard region. We call these the R and U models, respectively. Using these models we obtained hazard curves and expected PGA values for two probabilities of exceedance, at 2 and 10 percent in 50 years for events exceeding magnitudes $M_w \geq 5$. These probabilities correspond to return periods of 475 and 2475 years. For nomenclature purposes, we use the probabilities as a suffix on the model type. As an example, we refer to the results for 2 percent probability of exceeding an event magnitude $M_w$ 5 in 50 years using the five-regions model as the R2, whereas for the case of 10 percent probability of exceedance when using the uniform model we would use the tag name U10.

@@

Fig.~\ref{fig:pga} shows the expected mean PGA values obtained using the selected attenuation relationship as a measured of gravity (g) throughout the region of interest for the different models, R2, R10, U2 and U10. As can be expected, the highest PGA values are associated with the R2 and U2 models, and the lowest with the R10 and U10 models. In results for the R models shown Fig.~\ref{fig:pga}a, the highest levels of expected ground motion occur in the Kopeh Dagh region (near the border with Turkmenistan), and in a small area half-way between the city of Karaj and the Piranshahr fault, where a small concentration of events can be observed in Fig.~\ref{fig:catalog}b, as well as along the Piranshahr and Morvarid fault lines between the Zagros and Azerbaijan seismic regions (see Fig.~\ref{fig:selected}a for reference). The Azerbaijan seismic zone does not present equally high levels of expected ground motion. In the U models shown in Fig.~\ref{fig:pga}b, on the other hand, the highest levels of expected ground motion are observed precisely in the area corresponding to the Azerbaijan seismic region, especially in the vicinity of the North Tabriz Fault system (see Fig.~\ref{fig:selected} for reference).

@@

As it can be seen in Fig.~\ref{fig:pga.ru.std}, the combined (R,U) model follows more closely the distribution of the region's seismicity shown in Fig.~\ref{fig:catalog} as well as the major faults shown in Fig.~\ref{fig:selected}a, and thus we consider it to be a better assessment of the seismic hazard in the region. It displays well the contribution of the North Tabriz Fault system in the Azerbaijan region as well as the contribution of the Piranshahr and Morvarid faults in the Zagros region and that of the different faults in the Alborz and Kopeh Dagh regions. In Fig.~\ref{fig:pga.ru.std}, the highest PGAs are observed in the (R,U)2+STD model results, where the expected accelerations are of the order of 1.4 g to the northwest of city of Tabriz. These maximum values should, nonetheless, be interpreted with caution given the fact that predictions are based on rock-site assumptions and do not include any kind of site effects or other related conditions which may often contribute to mitigate peak accelerations.

@@

We also note that the overall distribution of mean PGA values shown in the middle panels in Fig.~\ref{fig:pga.ru.std} are the ones that best agree with the recent study by \citet{Khodaverdian_2016_BSSA}. These estimates also agree, on average, with other previous studies. For the north-west seismic regions of Azerbaijan and Zagros, for instance, \citet{Tavakoli1999} predicts maximum mean PGA (10\% in 50 years) values of 0.45 g near the North Tabriz Fault and the North Tehran Fault zones, where we find equivalent values to be between 0.35 and 0.5 g in the 10(R,U)5 model. Similarly, for the particular case of the city of Tabriz, \citet{Vafaie2011} obtained slightly higher expected accelerations on the order of 0.30--0.90 and 0.20--0.65 g for 2 and 10\%, respectively, where our mean values are about 0.53 and 0.36 g for the same corresponding probabilities. In the north-central region of Alborz, \citet{Ghodrati2003} and \citet{Boostan2015} predict PGA values of the order 0.27--0.46 g and 0.42--0.48 g for Tehran, where we found equivalent results to be about 0.25--0.35 g for a 10\% probability of exceedance. The latter estimate is in good agreement with \citet{Abdi2013}, who predicts PGA values in the vicinity of Karaj and Tehran to be on the order of 0.27--0.31 g for 10\% probability of exceedance. Results from \citet{Ghodrati2008} for the Gilan province around the city of Rasht are also on the same order of those shown in Fig.~\ref{fig:pga.ru.std}. Finally, for the city of Bojnurd in the northeast province of Khorasan in the Kopeh Dagh seismic region, \citet{Rahgozar2012} reports expected PGA values in the ranges of 0.28--0.49 and 0.15--0.22 g for 2 and 10\%, where we obtain mean values of 0.51 and 0.34 g, respectively.

Table \ref{tab:pga} shows a condensed version of these comparisons between our results and those of other previous studies for selected cities in northern Iran. As mentioned before, overall, our results are close or within the ranges from most studies. The only noticeable exceptions are the results of \citet{Golara2014}, whose expected PGA values are significantly higher the rest, especially in the cities of Tabriz, Karaj, Tehran and Mashhad, where their results are closer to those of our upper (+STD) range of values.

% ***

% Although not included here, we also computed expected PGA values for a magnitude threshold $M_w$ 4.5. In that case, our results are closer to those of \citet{Golara2014}, but deviate from the rest. In general, the smoothed seismicity method tends to lead to lower hazard estimations than those obtained using specific fault definitions (as opposed to point sources) where the concept of a threshold becomes unnecessary and the hazard, and thus is sensitive to the appropriate selection of the magnitude threshold.

% ***

Following the comparisons shown in Table \ref{tab:pga}, we looked into the more complete picture of the annual probabilities of exceedance as functions of PGA. The results are shown in Fig.~\ref{fig:hazardcurve} for six of the main cities in northern Iran considering the two basic R and U models and the combined (R,U) model. In each case we also show the range $\pm 1$ standard deviation (shaded areas). These hazard curves exemplify again the differences between the R and U models, which are more prominent in the northwest cities of Tabriz and Ardabil, where the U models yield higher expected values of PGA for any given annual probability of exceedance. Minor differences are also present in the results of the central cities of Karaj and Tehran, whereas for the cities of Urmia and Mashhad in the Zagros and Kopeh Dagh seismic zones, the R and U models yield similar if not equivalent hazard estimates. This is consistent with our previous observations regarding Figs.~\ref{fig:pga}, \ref{fig:pga.diff} and \ref{fig:pga.ru.std}.

@@

Last, we put additional attention on the results for the capital city of Tehran. We compare hazard curves obtained in this study against those computed by \citet{Ghodrati2003} and \citet{Boostan2015}. Our results are similar to those obtained by \citet{Ghodrati2003} in the range between 0.2 and 0.7 g (corresponding to annual probabilities of exceedance between $10^{-5}$ and $10^{-2}$), but the trend indicates we would obtain increasingly higher probabilities associated with PGA values above 0.8 g. Conversely, the trend in the hazard curve from \citet{Boostan2015} suggest the city is to expect significantly higher levels of ground motion at higher probability rates, which is contrary to our findings and those of \citet{Ghodrati2003}.
